---
title: "Estimating the Phillips Multiplier with R"
author: "Itamar Caspi"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: no
    theme: journal
    toc: yes
    toc_depth: 4
    toc_float: yes
abstract: |
  The following code replicates some of the results that appear in ["The Phillips Multiplier" by Barnichon and Mesters (2020)](https://www.sciencedirect.com/science/article/abs/pii/S0304393220300520). In short, The Phillips multiplier is a statistic to non-parametrically characterize the central bank inflation-unemployment trade-off.
---

__NOTE__: The following code cannot run simply as-is since it relies on sourcing several auxiliary functions. The easiest way to use this notebook is to clone the entire R project folder which is located [here](https://github.com/itamarcaspi/phillips-multiplier), open the project on Rstudio, and knit this R markdown notebook.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval    = TRUE,
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## Load Packages
```{r packages}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,  # for data wrangling and plotting
  recipes,    # for data preprocessing
  readxl,     # for reading xls, xlsx files
  lubridate,  # for dealing with dates
  hrbrthemes, # for a fancy ggplot theme
  latex2exp   # for math symbols in plot titles
)
```

## Source auxiliary functions
```{r, results='hide'}
paste0("functions/", list.files("functions")) %>% 
  map(source)
```

## Setup
```{r params}
data_source <- 1                      # which data to use? 1=Romer, 2=HF
endoLags    <- 4                      # no. of lags for endogenous variables
iH          <- 21                     # horizon for IRF
vParB       <- seq(-10, 5, by = 0.01) # grid of \beta parameters to test
CI          <- 0.95                   # confidence interval

set.seed(2805)                        # for replication
```

## Read data
```{r read}
if (data_source == 1) {
  df_raw <- read_xls("data/DataWeakIVpm.xls") 
} else if (data_source == 2) {
  df_raw <- read_xlsx("data/DataWeakIVhffinal.xlsx")
}
```

## Process data
```{r process}
if (data_source == 1) {
  df <- 
    df_raw %>%
    mutate(
      iv_m = iv_m1 + iv_m2 + iv_m3,
      pi_a = pi - piE,
      ugap = Upgap,
      date = as_date(date_decimal(year))
    )
}
if (data_source == 2) {
  df <- 
    df_raw %>%
    mutate(
      iv_m  = FF4star_G_m1 + FF4star_G_m2 + FF4star_G_m3,
      pi    = PI_pce,
      pi_a  = pi - PIx_pce,
      ugap  = UR_gap,
      date  = as_date(date_decimal(obs))
    )
}
```

## Select variables
```{r select_variables}
df_pi <- 
  df %>% 
  select(date, pi, ugap, iv_m)

df_pi_a <- 
  df %>% 
  select(date, pi_a, ugap, iv_m)
```

## Plot variables
```{r plot_vars}
df_pi %>% 
  pivot_longer(-date, names_to = "name", values_to = "value") %>% 
  ggplot(aes(date, value)) +
  geom_line(size = 0.5) +
  facet_wrap(~ name, scales = "free") +
  labs(
    x = "",
    y = "",
    subtitle = ""
  ) +
  theme_ipsum() +
  theme(
    axis.title.x=element_text(size=14),
    axis.title.y=element_text(size=12)
  )
```

## Generate lags
```{r lags}
df_lag <-
  recipe(~ ., data = df_pi) %>% 
  step_lag(pi, ugap, lag = 1:endoLags) %>% 
  prep(df_pi) %>% 
  bake(df_pi)

df_lag_a <-  
  recipe(~ ., data = df_pi_a) %>% 
  step_lag(pi_a, ugap, lag = 1:endoLags) %>% 
  prep(df_pi_a) %>% 
  bake(df_pi_a)
```

## Extract vectors and matrices
```{r vectors}
vPi  <- df_lag %>% pull(pi)
mX   <- df_lag %>% select(starts_with("lag_")) %>% as.matrix()

vPiA <- df_lag_a %>% pull(pi_a)
mXA  <- df_lag_a %>% select(starts_with("lag_")) %>% as.matrix()

vU   <- df_lag %>% pull(ugap)
mZ   <- df_lag %>% pull(iv_m)
```

## Estimate
```{r estimate}
results <- 
  PhillipsMultiplier(vParB, iH, vPi, vU, mZ, mX, CI) %>% 
  as_tibble()

results_a <- 
  PhillipsMultiplier(vParB, iH, vPiA, vU, mZ, mXA, CI) %>% 
  as_tibble()
```

## Plot $\mathcal{P}_{h}$
```{r pm}
p_pm <- 
  results %>% 
  ggplot(aes(h, vBeta)) +
  geom_line(size = 1, color = "black") +
  geom_ribbon(
    aes(ymin = vLowerAR, ymax = vUpperAR),
    fill = "#05827e",
    alpha = 0.2
  ) +
  geom_hline(aes(yintercept = 0), linetype = 2) +
  xlim(0, iH) +
  labs(
    x = "Quarters",
    y = TeX("$P_{h}$"),
    subtitle = ""
  ) +
  theme_ipsum() +
  theme(
    axis.title.x=element_text(size=14),
    axis.title.y=element_text(size=12)
  )

p_pm
```

## Plot unconditional PM
```{r upm}
p_pm_uc <- 
  results %>% 
  ggplot(aes(h, vBetauc)) +
  geom_line(size = 1, color = "black") +
  geom_ribbon(
    aes(ymin = vLowerARuc, ymax = vUpperARuc),
    fill = "#05827e",
    alpha = 0.2
  ) +
  geom_hline(aes(yintercept = 0), linetype = 2) +
  xlim(0, iH) +
  labs(
    x = "Quarters",
    y = TeX("$P_{h}$"),
    subtitle = ""
  ) +
  theme_ipsum() +
  theme(
    axis.title.x=element_text(size=14),
    axis.title.y=element_text(size=12)
  )

p_pm_uc
```

## Plot $F$-statistics
```{r fstat}
p_f <- 
  results %>% 
  ggplot(aes(h, vFstat)) +
  geom_col(fill = "#05827e") +
  theme_ipsum() +
  labs(
    x = "Quarters",
    y = TeX("F-stat")
  ) +
  theme(
    axis.title.x=element_text(size=14),
    axis.title.y=element_text(size=12)
  )

p_f
```

## Plot $\mathcal{R}_{h}^{\bar{u}}$
```{r irf_u}
irf_scale <- max(results$vBetaUr) #rescale IRFs so that ir of ur peaks at 1

p_ur <- 
  results %>% 
  ggplot(aes(h, vBetaUr/irf_scale)) +
  geom_line(size = 1, color = "black") +
  geom_ribbon(
    aes(ymin = vLowerUR/irf_scale, ymax = vUpperUR/irf_scale),
    fill = "#05827e",
    alpha = 0.2
  ) +
  geom_hline(aes(yintercept = 0), linetype = 2) +
  xlim(0, iH) +
  labs(
    x = "Quarters",
    y = TeX("$R_{h}^{\\bar{u}}$"),
    subtitle = ""
  ) +
  theme_ipsum() +
  theme(
    axis.title.x=element_text(size=14),
    axis.title.y=element_text(size=12)
  )

p_ur
```

## Plot $\mathcal{R}_{h}^{\bar{\pi}}$
```{r irf_pi}
p_pi <- 
  results %>% 
  ggplot(aes(h, vBetaPi/irf_scale)) +
  geom_line(size = 1, color = "black") +
  geom_ribbon(
    aes(ymin = vLowerPI/irf_scale, ymax = vUpperPI/irf_scale),
    fill = "#05827e",
    alpha = 0.2
  ) +
  geom_hline(aes(yintercept = 0), linetype = 2) +
  xlim(0, iH) +
  labs(
    x = "Quarters",
    y = TeX("$R_{h}^{\\bar{\\pi}}$"),
    subtitle = ""
  ) +
  theme_ipsum() +
  theme(
    axis.title.x=element_text(size=14),
    axis.title.y=element_text(size=12)
  )

p_pi
```

## Plot $\mathcal{K}_{h}$
```{r kappa}
p_k <- 
  results_a %>% 
  ggplot(aes(h, vBeta)) +
  geom_line(size = 1, color = "black") +
  geom_ribbon(
    aes(ymin = vLowerAR, ymax = vUpperAR),
    fill = "#05827e",
    alpha = 0.2
  ) +
  geom_hline(aes(yintercept = 0), linetype = 2) +
  xlim(0, iH) +
  labs(
    x = "Quarters",
    y = TeX("$\\kappa_{h}$"),
    subtitle = ""
  ) +
  theme_ipsum() +
  theme(
    axis.title.x=element_text(size=14),
    axis.title.y=element_text(size=12)
  )

p_k
```


## References
Barnichon, R., & Mesters, G. (2020). The Phillips Multiplier. _Journal of Monetary Economics_.